---
title: "Analysis of the Near Earth Object Asteroid Dataset"
output: 
  html_document:
    keep_md: true
---

#### Kory Becker - October 21, 2015

## Synopsis

This document is a basic analysis of the NASA [Near Earth Object](http://neo.jpl.nasa.gov/ca/) (NEO) database, with specific highlight of the October 31, 2015 asteroid, 2015 TB145.

The NEO database contains a list of asteroids and comets that have been determined to have a perihelion distance of less than 1.3 AU. Comets are further restricted to those within an orbital period of less than 200 years.

The analysis shows that the 2015 TB145 asteroid is a particular outlier in the current dataset, specifically with regard to velocity, size, and distance to Earth.

## Data Processing

The data source comes from the table of near Earth objects on the NASA JPL NEO Close Approach [page](http://neo.jpl.nasa.gov/ca/). The table of data was extracted and saved to a tab-separated [file](https://raw.githubusercontent.com/primaryobjects/asteroids/master/asteroids.tsv), which can be processsed, as shown below.

```{r}
library(stringr)

# Read data.
data <- read.csv('asteroids.tsv', sep='\t', stringsAsFactors = FALSE)

# Set column names.
names(data) <- c('name', 'date', 'au', 'ld', 'diameter', 'h', 'velocity')

# Cleanup name, remove parenthesis.
data$name <- gsub('[\\(\\)]', '', data$name)

# Calculate average size, based on low and high range on diameter.
data$size <- sapply(data$diameter, function(d) {
  match <- str_match(d, '([-+]?[0-9]*\\.?[0-9]+) (k?m)( - ([-+]?[0-9]*\\.?[0-9]+) (k?m))?')

  # Extract low value.
  low <- as.numeric(match[2])
  if (match[3] == 'km') {
    # Convert to meters.
    low <- low * 1000
  }
  
  # Extract high value, if available.
  high <- low
  if (!is.na(match[5])) {
    high <- as.numeric(match[5])
    
    if (match[6] == 'km') {
      # Convert to meters.
      high <- high * 1000
    }
  }
  
  # Use average of low and high.
  mean(c(low, high))
})
```

Since the diameter of objects in the table coorespond to a range of values, we take the mean value of the low and high range to calculate an average size.

## Size vs Velocity

We can visualize the data with regard to diameter and velocity by plotting the objects. The immediate outliers in the data become apparent in the resulting graph. Two objects, in particular, represent the most extreme ends of the graph, They include [2015 TB145](http://ssd.jpl.nasa.gov/sbdb.cgi?sstr=2015%20TB145;orb=1) and [163696 (2003 EB50)](http://ssd.jpl.nasa.gov/sbdb.cgi?sstr=163696;orb=1).

2015 TB145 has the highest velocity in the dataset, at 34.95 km/s. 163696 (2003 EB50) has the largest size at a mean of 2,250m in diameter.

```{r, echo=FALSE}
# Plot size and velocity.
plot(velocity ~ size, data = data, xlab = 'Average Diameter (m)', ylab = 'Velocity', col=ifelse(name == '2015 TB145', 'red', 'black'), pch=ifelse(name == '2015 TB145', 19, 1), cex=ifelse(name == '2015 TB145', 2, 1))
with(data, text(velocity ~ size, labels = data$name, cex=0.5, pos=4))
# Draw trend-line.
fit <- lm(velocity ~ size, data)
abline(fit)
```

*Figure 1a. Visualizing the near Earth object database as velocity by diameter. The asteroid 2015 TB145 represents the fastest moving object in the dataset at 34.95 kb/s.*

```{r, echo=FALSE}
# Order data by velocity, descending.
head(data[order(-data$velocity),c(1:3, 5, 7)])
```

## Distance to Earth by Size

We can visualize one of the most critical points with respect to Earth, by displaying the distance of each object by its cooresponding size. The asteroid 2015 TB145 is the closest approaching object in the dataset. The asteroid will pass by Earth at an estimated distance of 0.0033 AU or 1.3 lunar distance.

In addition to its estimated approach proximity to Earth, 2015 TB145 is classified by NASA JPL with a [condition code](http://ssd.jpl.nasa.gov/sbdb_help.cgi?name=condition_code) 8. Condition code is defined as, "orbit uncertainty estimate 0-9, with 0 being good, and 9 being highly uncertain".

```{r, echo=FALSE}
# Plot distance and size.
plot(au ~ size, data = data, xlab = 'Average Diameter (m)', ylab = 'Distance to Earth (au)', col=ifelse(name == '2015 TB145', 'red', 'black'), pch=ifelse(name == '2015 TB145', 19, 1), cex=ifelse(name == '2015 TB145', 2, 1))
with(data, text(au ~ size, labels = data$name, cex=0.5, pos=4))
# Draw trend-line.
fit <- lm(au ~ size, data)
abline(fit)
```

*Figure 2a. The asteroid 2015 TB145, highlighted in red, represents the closest approaching object to Earth included in the dataset.*

```{r, echo=FALSE}
# Order data by distance, increasing
head(data[order(data$au),c(1:3, 5, 7)])
```


## Distance to Earth by Velocity

The velocity of 2015 TB145 also displays as an outlier, with regard to the dataset. The asteroid has a velocity of 34.95 km/s and is the fastest moving near Earth object within the dataset.

```{r, echo=FALSE}
# Plot distance and velocity
plot(au ~ velocity, data = data, xlab = 'Velocity', ylab = 'Distance to Earth (au)', col=ifelse(name == '2015 TB145', 'red', 'black'), pch=ifelse(name == '2015 TB145', 19, 1), cex=ifelse(name == '2015 TB145', 2, 1))
with(data, text(au ~ velocity, labels = data$name, cex=0.5, pos=4))
# Draw trend-line.
fit <- lm(au ~ velocity, data)
abline(fit)
```

*Figure 3. The asteroid 2015 TB145, highlighted in red, represents the fastest moving and closest approaching object within the dataset.*

## Conclusion

Through the analysis of the NASA JPL Near Earth Object database, we've determined that currently, the object with the highest velocity is the 2015 TB145 asteroid, due for fly-by on October, 31 2015. This asteroid is also the largest closest passing object currently in the active dataset. A second outlier object is 163696 (2003 EB50), holding the largest diameter of all objects in the dataset, due for fly-by on November 29, 2015. Of important note is the condition code 8 classification for 2015 TB145, indicating the asteroid's orbit is highly uncertain. This may be a result of the recent discovery of its presence, combined with the high velocity of movement.

## Sources

1. NASA, Near Earth Object Program *[[link]](http://neo.jpl.nasa.gov/index.html)*

2. NASA, Near Earth Object Program, Close Approaches *[[link]](http://neo.jpl.nasa.gov/ca/)*

3. JPL Small-Body Database Browser, 2015 TB145 Orbit Determination Parameters *[[link]](http://ssd.jpl.nasa.gov/sbdb.cgi?sstr=2015%20TB145;orb=1)*
